\name: A Non-Parametric, Hierarchical \\ Wavelength-Calibration Model for a Precision Spectrograph

### 2020-05-14 paper/paper.tex
Proofread and New Figures

\begin{abstract}\noindent%
\name\ is a hierarchical, non-parametric framework for wavelength calibrating precision spectrographs.  This method capitalizes on recent hardware improvements in highly-stabilized, extreme-precision radial-velocity spectrographs.  With stabilized spectrographs, the full optical system and detectors can only vary within a tiny range of configurations.  This allows us to use all calibration data to determine the entire space of possible calibrations for a stabilized spectrograph.  A wavelength solution then need only pinpoint where in this low-dimensional calibration space the spectrograph exists.  \name\ also takes advantage of the development of laser-frequency combs or etalons, which generate a dense set of stable calibration points.  This density of points permits the freedom of a non-parametric wavelength solution that can match or adapt to any instrument or detector oddities, more so than any one polynomial.  We demonstrate the success of this method with data from the EXtreme PREcision Spectrograph, \expres, which includes a Menlo Systems laser frequency comb.  We find that \name\ returns more accurate wavelength predictions by more than a factor of 2 over classic, non-hierarchical, polynomial solutions.  Employing \name\ wavelengths on \expres\ RV measurements on HD 3651 reduced the RMS of a planet fit by \lz{blah}.
\end{abstract}

@@

\section{Introduction} 
Extremely precise, radial-velocity (\eprv) programs have been incredibly fruitful in finding and characterizing extra-solar planets \citep{mayor2011, bonfils2013, plavchan2015, butler2017}.  These programs typically make use of spectrographs with resolutions on the order of $10^5$, which correspond to line widths on the order of $3000\,\mps$.  Exoplanet science happens at the level of $3\,\mps$ precision, with current instruments aiming to reach $0.1\,\mps$ precision \citep{fischer2016}.  This requires new spectrographs to be calibrated or stabilized to better than $10^{-4}$ of a pixel (assuming that the spectrographs are well sampled).    Some hardware--software systems, such as \espresso, \expres, and \neid\ were designed to deliver precision at this level \citep{pepe2013,  jurgenson2016, neid}.  Early data from \expres\ and \espresso\ suggest that the hardware improvements do propagate through to sub-$\mps$ uncertainties on final, on-sky RV measurements \citep{blackman2020, petersburg2020, pepe2020}.

@@

Here we propose to simplify and improve calibration programs for \eprv\ hardware systems with two very simple but innovative ideas.  The first flows from the fact that calibration sources---which include arc lamps (in some wavelength ranges), etalons, and laser-frequency combs (\lfc s)---illuminate the spectrograph with very stable, very dense sets of lines; almost every location in the spectrograph image plane is surrounded by nearby, useful calibration lines.

This recommends a calibration methodology that is \emph{non-parametric}:  If every point in the spectrograph detector is surrounded by nearby calibration lines, the wavelength solution can, for example, be made simply as an interpolation of the calibration data.  The density of lines removes the need to enforce any functional form for the wavelength solution (such as a continuous eighth-order polynomial, for example).  In some ways, this is a generalization of the recent work that demonstrated the efficacy of constructing a wavelength solution as multiple, segmented polynomials \citep{milakovic2020}.  Going non-parametric will improve calibration accuracy by not forcing the choice of a parametric form that may bias the calibration, especially when the chosen function is inappropriate (as, for example, polynomials will be at detector edges).

The second simple idea follows from the observation that contemporary \eprv\ instruments are incredibly stable.  Temperature-controlled, fiber-fed spectrographs vary only slightly over the night or season.  This suggests that such stabilized spectrographs vary along only a small number of axes in what you might call ``calibration space'', or the (very high dimensional) space of all possible wavelength solutions.  That is, not only are the spectrographs stable, but they also have few environmentally accessible degrees of freedom.  This renders it inadvisable to fit each calibration exposure or calibrate each science exposure independently.  Instead, all the calibration data (or all the data) should be used to determine the space in which the instrument can and does vary.  Subsequent calibration work then need only determine where in the accessible part of calibration space the spectrograph was located for each exposure.

In the context of probabilistic models, this structure is \emph{hierarchical}:  The calibration data are used not just to determine the wavelength solution at one moment, but also to determine the possible \emph{space} of wavelength solutions at all moments.  In the statistics literature this concept is often described as \emph{de-noising}:  Every calibration exposure contains information about every other exposure.  Thus every exposure can be improved (i.e. de-noised) with information from every other exposure.

The method we propose here---\name---embodies these ideas.
It is a non-parametric, hierarchical, data-driven model for the wavelength solution.  By being non-parametric, it delivers enormous freedom to the wavelength solution to match or adapt to any instrument or detector oddities.  By being hierarchical, it restricts that freedom tremendously, but it does so appropriately for the empirically determined variations in the stabilized spectrograph.  \name\ is designed for temperature-controlled, fiber-fed spectrographs with good calibration sources, such as laser-frequency combs, or etalons.  We have in mind \eprv\ instruments and \eprv\ science cases, but we expect \name\ to have applications for other kinds of spectrographs in other contexts.  \name\ should be applicable to almost every astronomical spectrograph.

\section{Method} \label{sec:method}
The \name\ method is designed to turn a series of calibration lines with known wavelengths and well-fit detector positions, and de-noise and interpolate them into a full wavelength model applicable to all exposures taken with the instrument.  \name\ operates on the two core ideas that the wavelength solution should be given enormous flexibility, but that it lives in a very low-dimensional space, where the degrees of freedom are set by the limited kinematics of the stabilized spectrograph hardware.

\name\ assumes dense enough calibration line coverage with well-fit line centers to provide some constraint on an interpolated wavelength solution across an echelle order.  Additionally, \name\ assumes the instrument hardware is stabilized enough that the space of possible calibration states is low-dimensional.  We revisit and quantify these conditions in Section \ref{sec:discussion}.

Wavelength calibration is usually posed in the following way.  Given an exposure $n$, and echelle order $m$, there is a relationship between
the two-dimensional $(x,y)$-position on the detector and the
wavelength $\lambda$
\begin{equation}
\lambda(x,y,m,n) = f(x,y,m;\theta_{n})
\quad ,
\label{eq:wsol}
\end{equation}
where $\theta_{n}$ represents the parameters describing a given exposure.

Classically, pipelines employ polynomials to construct smooth wavelength solutions for each exposure.  For example, the \expres\ pipeline sets the function $f(x,m;\theta_{n})$ from Equation \ref{eq:wsol} to be a 2D, 9\textsuperscript{th}-order polynomial where $\theta_{n}$ represents the polynomial coefficients, $c_{nij}$, unique to each exposure $n$ \citep{petersburg2020}.
\begin{equation}
\lambda(x,m,n) = \sum_{i=0}^9\sum_{j=0}^9 c_{nij}\, x^i\,m^j + \mathrm{noise}
\quad ,
\label{eq:poly_wsol}
\end{equation}
Here the y-dependence is dropped as that dependence is carried by spectral order $m$.  The line position information is therefore uniquely identified by echelle order $m$ and pixel in the dispersion direction, $x$.  The coefficients $c_{nij}$ are interpolated to time $t_n$ of exposure $n$ by a third-order polynomial of time.  This third-order polynomial is evaluated at the time of non-calibration exposures to re-construct a 2D, 9th-order polynomial wavelength solution for that exposure.  Each calibration image obtains its $c_{nij}$ coefficients independently.

@@

The challenge is to learn these basis functions from the data and get the $K$ amplitudes, $a_{nk}$, for every exposure $n$.  There are many ways to discern these basis functions.  In this paper, we present a model using principal component analysis (PCA).  A PCA is justifiable in the limit where there is very high signal-to-noise ratio, as is usually the case with typical calibration images.  There are many alternatives to PCA for this dimensionality reduction; we return to this point in the discussion (\textsection \ref{sec:discussion}) below.

\begin{figure*}[h]
\centering
\includegraphics[width=\textwidth]{Figures/methodCartoon.pdf}
\caption{A cartoon representation of the \name\ method, as described in Section \ref{sec:method}.  Variations in line position, amplitude of basis vectors, and interpolation greatly enlarged to show texture.  In step one, dimensionality reduction and denoising, the complete set of line positions for each exposure is analyzed to return a set of basis vectors, $g_n$, which span the calibration space of the spectrograph.  In step two, \name\ interpolates the amplitudes of each basis vector, $a_n,k$, to return the calibration state for a specific exposure, given as a set of de-noised lines.  The assigned wavelengths of these de-noised line positions are then interpolated onto other pixels in step three to construct a full wavelength model.}
\label{fig:testHists}
\end{figure*}

@@

\subsection{?Interpolating Calibration Position?}
 \label{sec:interp_time}
The magnitude of each principal component can be interpolated with respect to housekeeping data, such as time, temperature, telescope position, etc. to determine different calibration states of the spectrograph.  The choice of what to interpolate against depends on the dominant contribution to variation in the instrument.

In the implementation of \name\ presented here, the magnitudes of the principal components are interpolated with respect to time.  New magnitudes $a_n'k$ are found for the midpoint time of non-calibration exposures and used to construct the calibration state of the spectrograph for that exposure.  Using interpolated magnitudes, $a_{n;k}$, and the basis vectors, $g_k(\lambda,m)$, returned by the de-noising process, \name\ can construct a set of calibration lines, $x(\lambda,m,n') $, for any exposure as prescribed in Equation \ref{eq:excl_wsol}.  The best interpolation scheme for \expres\ is discussed below in Section \ref{sec:choice_avt}, but may vary for different instruments.

@@

Find $a_{n',k}$ by interpolating $a_{n,k}$ with to $t_{n'}$\;
$x(\lambda,m,n') = g_0(\lambda,m) + \sum_{k=1}^K a_{n'k}\,g_k(\lambda,m)$ where $K=6$\;
\For{each unique $m$}{
	interpolate $\lambda$ with respect to $x(\lambda,m,n')$ onto pixels $x'(m,n')$\;
	}
\caption{Generating Wavelength Solution}
\end{algorithm}

@@

Rather than using a simultaneous calibration fiber, two to three LFC exposures are interspersed between science exposures every 15-30 minutes.  ThAr exposures are taken at the beginning and end of each night.  All calibration data is taken through the science fiber, meaning they traveling down the same optical pathway and are projected onto the same pixels as all science observations.

LFC lines cover 40 echelle orders, which contain 19203 lines.  Though the results are primarily based on work with LFC data, there is some discussion of applications to arc lamps.   ThAr lines cover all 86 extracted orders of \expres, which includes 5295 lines.

For both the LFC and ThAr data, lines that appeared in less than 60\% of exposures were not included in the analysis.  Similarly, files with more than 60\% of expected lines missing were cut from the analysis.

@@

In order to satisfy the assumption of stability needed for \name, only exposures from only one ``instrumental epoch" of \expres\ are used.   \expres\ exposures are separated into different instrumental epochs that correspond to atypical hardware changes--changes that altered the position of the echellogram on the detector, the shape of the instrumental PSF, or the calibration sources.  Significant instrumental changes break the assumption of a stabilized instrument with only low-dimensional variations.  Each instrument epoch is therefore treated independently.  In the results presented here, we use 1227 LFC exposures and 78 ThAr exposures taken between October 14 and December 18, 2020 on 29 unique nights.

\section{Tests}\label{sec:tests}
We perform a series of tests to validate the performance of \name\ and benchmark \name -generated wavelengths against wavelengths generated by a classic, parametric method.  For all these tests, we leave out a subset of calibration lines with known wavelengths as a ``validation'' sample, generate wavelengths for these lines, and compare the predicted wavelength to the assigned wavelength of each line.  This inherently folds in errors in the measured line center of each calibration line, but this contribution to the residuals will be the same across all tests.

@@

\begin{figure}[h]
\centering
\includegraphics[width=.5\textwidth]{Figures/all_results.pdf}
\caption{Difference in predicted and assigned wavelengths for the wavelength calibration tests described in Section \ref{sec:tests}.  The standard deviation of the residuals for each test is given in the top-right corner in each method's corresponding color.}
\label{fig:testHists}
\end{figure}

@@

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{Figures/lineResids2D.png}
%\includegraphics[width=\textwidth]{Figures/lineResids2D.pdf}
\caption{Residuals for a single image plotted with respect to detector position (as defined by echelle order and x-pixel) for various wavelength calibration methods.  Each line is colored by the difference between the predicted wavelength and the true wavelength for each line, given in units of $/mps$.  Residuals to a classic, polynomial fit is shown in the left-most plot.  The center plot shows residuals to an interpolated wavelength solution.  The right-most plot shows residuals when a ThAr wavelength solution is interpolated onto LFC line positions.  The position of ThAr lines are over-plotted in black.}
\label{fig:resid2d}
\end{figure*}

\begin{figure}[t]
\centering
\includegraphics[width=0.5\textwidth]{Figures/lineDensity.png}
\caption{Comparison of polynomial-fit and interpolated wavelength solutions using either just ThAr lines or LFC lines.  For a subset of one order, the difference between the assigned and model wavelengths for each LFC line is plotted with respect to pixel.  The large residuals between two widely-spaced ThAr lines when implementing an interpolated wavelength solution with just ThAr lines (blue curve) extends out of frame.  The classic, polynomial fit exhibits similar residuals in both amplitude and shape regardless of whether the set of ThAr lines or LFC lines are used.  An interpolated wavelength solution using LFC lines (orange curve) exhibits the lowest residuals.}
\label{fig:waveResids}
\end{figure}

The residuals of the classic, polynomial wavelength solution is shown in the left-most plot of Figure \ref{fig:resid2d}.  There is a lot of vertical structure and some hints of a periodic, diagonal structure as well.  The residuals of the interpolation test for the same exposure is shown in the center plot of Figure \ref{fig:resid2d}.  There is no coherent structure here and smaller residuals.  We believe the flexibility of the interpolated model was able to account for high-order detector defects, which emerged as structure in the residuals of the classic, smooth, polynomial-driven wavelength solution.  In the \expres\ data, for example, the interpolated model accounted for \lz{detector effect STA will explain to us}.

The right-most plot of Figure \ref{fig:resid2d} shows the residuals when ThAr exposures, which have much fewer lines than LFC exposures, are run through \name\ and used to predict wavelengths for the completely independent LFC exposures taken during the same range of time.  Over-plotted in black are the positions of the ThAr lines.

Though the interpolated wavelength solution returns lower, less-structured residuals than the polynomial wavelength solution when guided by LFC lines, we see in the right-most plot of Figure \ref{fig:resid2d} that an interpolated wavelength solution can return much worse results in the space between widely separated calibration lines.

The move to an interpolated wavelength solution is driven by the assumption that a high density of calibration lines allows for more freedom in the resultant wavelength solution.  Introducing this flexibility is no longer justified in the regime where there are large spaces between calibration lines, as is the case with a classic ThAr lamp.

@@

\subsection{Impact on Radial Velocity Measurements}\label{sec:test-rv}
\lz{Debra and RP currently working through this.  Results seem good.  Will ask them at next group meeting what we should include here}

\section{Choose Your Own Implementation} \label{sec:choices}
We describe here only one implementation of \name.  Using PCA and an interpolated wavelength solution is a fairly trivial step towards a complete hierarchical and non-parametric wavelength model.  It is possible to upgrade both the denoising and wavelength solution steps to true models.

@@

The wavelength solution could also move past interpolation.  For example, a Gaussian process could be used that is constrained to ensure monotonicity.  Replacing each step with a model will allow for full, hierarchical Bayesian inference.  This means error from wavelength calibration could be marginalized out.  This is only worth doing, however, if wavelength calibration is a significant fraction of the error budget.

The implementation of \name\ presented here, using PCA for denoising and interpolating a wavelength solution, incorporated various global variables and methods that we believe are or are close to optimal for constructing a high-fidelity wavelength solution.  The following subsections will describe each choice and the associated decision-making process.

@@

We settled on a $K$ value of 6.  Figure \ref{fig:pcLfc} shows the first 6 principal components constructed using LFC lines.  In both cases, there is clear structure in the first and second principal components.  The diagonal-banding structure seen in components 3, 5, and 6 we believe to be capturing detector fabrication defects.  Components 3 and 4 show aberrant behavior on the edges of the two bluest echelle orders, where lower signal results in more variation in the line fits.

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{Figures/pcsLfc6.png}
%\includegraphics[width=\textwidth]{Figures/pcsLfc6.pdf}
\caption{The first six principal components constructed using LFC lines.  Each line is plotted according to its echelle order and x-pixel and colored by the value of the principal component for that line.}
\label{fig:pcLfc}
\end{figure*}

In deciding a K value, we ran denoising tests, as described in Section \ref{sec:tests}, for K values spanning from 2 to 512.  One would expect the returned wavelengths to get increasingly more accurate with larger K until components that represent only noise are incorporated.  We found that the returned wavelengths plateau at a K value of 32 (see Figure \ref{fig:kvals}).  However, the improvement is slight with K values greater than six.  Comparisons of wavelengths returned by a K=6 model vs. a K=32 model show significant differences for less than 10 bluer lines, which are know to have greater error in their measured line positions.  A visual inspection of the principal components, such as the one displayed in Figure \ref{fig:pcLfc}, additionally show most of the structure is being captured in the first six principal components with later components dominated by noise.

@@

The choice in interpolation method can help guide how many wavelength calibration images are truly needed.  It is unnecessary to take calibration images at times where similar information can be filled in by a well-chosen interpolation scheme.  For example, with the \expres\ data shown here, it is clear that nightly calibration images are needed, but for a linear trend, only two calibration images throughout the night are strictly required.

The amplitudes $a_{n,k}$ can also be interpolated with respect to any housekeeping data, not just time.  Best results should come from interpolating with respect to whichever data is most strongly correlated with the calibration state of the spectrograph.  For example, with \expres, which is passively temperature controlled, the returned amplitudes $a_{n,k}$ were extremely correlated with the enclosure temperature, suggesting it would also be possible to interpolate with respect to temperature.

>>>>

\begin{abstract}\noindent%
\name\ is a hierarchical, non-parametric framework for wavelength calibrating spectrographs.  This method capitalizes on recent hardware improvements in highly-stabilized, extreme-precision radial-velocity spectrographs.  With stabilized spectrographs, the full optical system and detectors can only vary within a tiny range of configurations.  This allows us to use all calibration data to determine the entire space of possible calibrations for a stabilized spectrograph.  A wavelength solution then need only pinpoint where in this low-dimensional calibration space the spectrograph exists.  \name\ also takes advantage of the development of laser-frequency combs or etalons, which generate a dense set of stable calibration points.  This density of points permits the freedom of a non-parametric wavelength solution that can match or adapt to any instrument or detector oddities, more so than any parametric model.  We demonstrate the success of this method with data from the EXtreme PREcision Spectrograph, \expres, which includes a Menlo Systems laser frequency comb.  We find that \name\ returns more accurate wavelengths by greater than a factor of 2 over classic, non-hierarchical, polynomial solutions.  Employing \name\ wavelengths on \expres\ RV measurements of HD 3651 reduced the RMS of a planet fit by \lz{blah}.
\end{abstract}

@@

\section{Introduction} 
Extremely precise, radial-velocity (\eprv) programs have been incredibly fruitful in finding and characterizing extra-solar planets \citep{mayor2011, bonfils2013, plavchan2015, butler2017}.  These programs typically make use of spectrographs with resolutions on the order of $10^5$, which correspond to line widths on the order of $3000\,\mps$.  Exoplanet science happens at the level of $3\,\mps$ precision, with current instruments aiming to reach $0.1\,\mps$ precision \citep{fischer2016}.  This requires new spectrographs to be calibrated or stabilized to better than $10^{-4}$ of a pixel (assuming that the spectrographs are well sampled).    Some hardware--software systems, such as \espresso, \expres, and \neid\ were designed to deliver precision at this level \citep{pepe2013,  jurgenson2016, neid}.  Early data from \expres\ and \espresso\ suggest that the hardware improvements truly propagate through to sub-$\mps$ uncertainties on final, on-sky RV measurements \citep{blackman2020, petersburg2020, pepe2020}.

@@

Here, we propose to simplify and improve calibration programs for \eprv\ hardware systems with two very simple but innovative ideas.  The first flows from the fact that calibration sources---which include arc lamps (in some wavelength ranges), etalons, and laser-frequency combs (\lfc s)---illuminate the spectrograph with very stable, very dense sets of lines; almost every location in the spectrograph image plane is surrounded by nearby, useful calibration lines.

This recommends a calibration methodology that is \emph{non-parametric}:  If every point in the spectrograph detector is surrounded by nearby calibration lines, the wavelength solution can, for example, be made simply as an interpolation of the calibration data.  The density of lines removes the need to enforce any functional form for the wavelength solution (such as a continuous ninth-order polynomial, for example).  In some ways, this is a generalization of the recent work that demonstrated the efficacy of constructing a wavelength solution as multiple, segmented polynomials \citep{milakovic2020}.  Going non-parametric will improve calibration accuracy by not forcing the choice of a parametric form that may bias the calibration, especially when the chosen function is inappropriate (as polynomials are at detector edges).

The second simple idea follows from the observation that contemporary \eprv\ instruments are incredibly stable.  Temperature-controlled, fiber-fed spectrographs vary only slightly over the night or season.  This suggests that such stabilized spectrographs vary along only a small number of axes in what you might call ``calibration space'', or the (very high-dimensional) space of all possible wavelength solutions.  That is, not only are the spectrographs stable, but they also have few environmentally accessible degrees of freedom.  This renders it inadvisable to fit each calibration exposure or calibrate each science exposure independently.  Instead, all the calibration data (or all the data) should be used to determine the space in which the instrument can and does vary.  Subsequent calibration work then need only determine where in the accessible part of calibration space the spectrograph was located for each exposure.

In the context of probabilistic models, this structure is \emph{hierarchical}:  The calibration data are used not just to determine the wavelength solution at one moment, but also to determine the possible \emph{space} of wavelength solutions at all moments.  In the statistics literature, this concept is often described as \emph{de-noising}:  Every calibration exposure contains information about every other exposure.  Thus every exposure can be improved (i.e. de-noised) with information from every other exposure.

The method we propose here---\name---embodies these ideas.
It is a non-parametric, hierarchical, data-driven method to generate a wavelength model.  By being non-parametric, it delivers enormous freedom to the wavelength solution to match or adapt to any instrument or detector oddities.  By being hierarchical, it restricts that freedom tremendously, but it does so appropriately for the empirically determined variations in the stabilized spectrograph.  \name\ is designed for temperature-controlled, fiber-fed spectrographs with good calibration sources, such as laser-frequency combs, or etalons.  We have in mind \eprv\ instruments and \eprv\ science cases primarily because the need for good wavelength calibration in this field is so great.  Irregardless, we expect \name\ to have applications for other kinds of spectrographs in other contexts.  \name\ should be applicable to almost every astronomical spectrograph.

\section{Method} \label{sec:method}
The \name\ method is designed to take a series of calibration lines with known wavelengths and well-fit detector positions, and de-noise and interpolate them into a full wavelength model applicable to all exposures taken with the instrument.  \name\ operates on the two core ideas that the wavelength solution should be given enormous flexibility, but that it lives in a very low-dimensional space, where the degrees of freedom are set by the limited kinematics of the stabilized spectrograph hardware.

\name\ assumes dense enough calibration line coverage with well-fit line centers to provide some constraint on an interpolated wavelength solution across an echelle order.  Additionally, \name\ assumes the instrument hardware is stabilized enough that the space of possible calibration states is low-dimensional.  We revisit and quantify these conditions in Section \ref{sec:others}.

Wavelength calibration is usually posed in the following way.  Given an exposure $n$, and echelle order $m$, there is a relationship between
the two-dimensional $(x,y)$-position on the detector and the
wavelength $\lambda$
\begin{equation}
\lambda(x,y,m,n) = f(x,y,m;\theta_{n})
\quad ,
\label{eq:wsol}
\end{equation}
where $\theta_{n}$ represents the parameters describing the wavelength solution for a given exposure.

Classically, pipelines employ polynomials to construct smooth wavelength solutions for each exposure.  For example, the \expres\ pipeline sets the function $f(x,y,m;\theta_{n})$ from Equation \ref{eq:wsol} to be a 2D, 9\textsuperscript{th}-order polynomial where $\theta_{n}$ represents the polynomial coefficients, $c_{nij}$, unique to each exposure $n$ \citep{petersburg2020}.
\begin{equation}
\lambda(x,m,n) = \sum_{i=0}^9\sum_{j=0}^9 c_{nij}\, x^i\,m^j + \mathrm{noise}
\quad ,
\label{eq:poly_wsol}
\end{equation}
Here the y-dependence is dropped as that dependence is carried by spectral order $m$.  The line position information is therefore uniquely identified by echelle order $m$ and pixel in the dispersion direction, $x$.  The coefficients $c_{nij}$ are interpolated to time $t_n$ of exposure $n$ by a third-order polynomial of time.  This third-order polynomial is evaluated at the time of non-calibration exposures to re-construct a 2D, 9th-order polynomial wavelength solution for that exposure.  Each calibration image obtains its $c_{nij}$ coefficients independently.

@@

The challenge is to learn these basis functions from the data and get the $K$ amplitudes, $a_{nk}$, for every exposure $n$.  There are many ways to discern these basis functions.  In this paper, we present a model using principal component analysis (PCA).  A PCA is justifiable in the limit where there is very high signal-to-noise ratio, as is usually the case with typical calibration images.  There are many alternatives to PCA for this dimensionality reduction; we return to this point in Section \ref{sec:choices} below.

\begin{figure*}[h]
\centering
\includegraphics[width=\textwidth]{Figures/methodCartoon.pdf}
\caption{A cartoon representation of the \name\ method, as described in Section \ref{sec:method}.  We enlarge variations in measured line position, changes in calibration space, and interpolation deviations for clarity.  In step one, dimensionality reduction and denoising, the complete set of line positions for all exposures is analyzed to return a set of basis vectors, $g_n$, which represent different ways the spectrograph calibration changes, and spans the possible calibration space of the spectrograph.  In step two, \name\ interpolates the amplitudes of each basis vector, $a_n,k$, to return the calibration state for a specific exposure, returned as a set of de-noised calibration lines.  The assigned wavelengths of these de-noised line positions are then interpolated onto other pixels in step three to construct a full wavelength model.}
\label{fig:testHists}
\end{figure*}

@@

\subsection{Interpolating Calibration Position}
 \label{sec:interp_time}
The magnitude of each principal component can be interpolated with respect to housekeeping data, such as time, temperature, telescope position, etc. to determine different calibration states of the spectrograph.  The choice of what to interpolate against depends on the dominant contribution to variation in the instrument.

In the implementation of \name\ presented here, the magnitudes of the principal components are interpolated linearly with respect to time.  \name\ finds new magnitudes $a_{n'k}$ for an exposure $n'$ at time $t_{n'}$, which are used to construct the calibration state of the spectrograph for that exposure.  Using interpolated magnitudes, $a_{n,k}$, and the basis vectors, $g_k(\lambda,m)$, returned by the de-noising process, \name\ can construct a set of calibration lines, $x(\lambda,m,n') $, for any exposure as prescribed in Equation \ref{eq:excl_wsol}.

@@

Find $a_{n',k}$ by interpolating $a_{n,k}$ with respect to $t_{n'}$\;
$x(\lambda,m,n') = g_0(\lambda,m) + \sum_{k=1}^K a_{n'k}\,g_k(\lambda,m)$ where $K=6$\;
\For{each unique $m$}{
	interpolate $\lambda$ with respect to $x(\lambda,m,n')$ onto pixels $x'(m,n')$\;
	}
\caption{Generating Wavelength Solution}
\end{algorithm}

@@

Rather than using a simultaneous calibration fiber, two to three LFC exposures are interspersed between science exposures every 15-30 minutes.  ThAr exposures are taken at the beginning and end of each night.  All calibration data is taken through the science fiber, so that they down the same optical pathway and are projected onto the same pixels as all science observations.

LFC lines cover 40 echelle orders, which contain 19203 lines.  Though the results are primarily based on work with LFC data, there is some discussion of applications to arc lamps.   ThAr lines cover all 86 extracted orders of \expres, which includes 5295 lines.  For both the LFC and ThAr data, lines that appeared in less than 60\% of exposures were not included in the analysis.  Similarly, files with more than 60\% of expected lines missing were cut from the analysis.

@@

In order to satisfy the assumption of only low-order variation needed for \name, we used exposures from only one ``instrumental epoch" of \expres.  \expres\ exposures are separated into different instrumental epochs that correspond to atypical hardware changes---changes that altered the position of the echellogram on the detector, the shape of the instrumental PSF, or the calibration sources.  Significant instrumental changes break the assumption of a stabilized instrument with only low-dimensional variations.  Each instrument epoch is therefore treated independently.  In the results presented here, we use 1227 LFC exposures and 78 ThAr exposures taken between October 14 and December 18, 2020 on 29 unique nights.

\section{Tests}\label{sec:tests}
We perform a series of tests to validate the performance of \name\ and benchmark \name -generated wavelengths against wavelengths generated by a classic, parametric method.  For all tests, we leave out a subset of calibration lines with known wavelengths as a ``validation'' sample, generate wavelengths for these lines using the remaining data, and compare the predicted wavelength to the assigned wavelength of each line.  This inherently folds in errors in the measured line center of each calibration line, but this contribution to the residuals will be the same across all tests.

@@

\begin{figure}[h]
\centering
\includegraphics[width=.5\textwidth]{Figures/all_results.pdf}
\caption{Difference in predicted and theoretical wavelengths for the wavelength calibration tests described in Section \ref{sec:tests}.  The standard deviation of the residuals for each test is given in the top-right corner in each method's corresponding color.}
\label{fig:testHists}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.5\textwidth]{Figures/lineResids2D_col.png}
%\includegraphics[width=0.5\textwidth]{Figures/lineResids2D_col.pdf}
\caption{Residuals for a single LFC exposure plotted with respect to detector position (as defined by echelle order and x-pixel) for various wavelength calibration methods.  Each line is colored by the difference between the predicted wavelength and the theoretical wavelength for each line, given in units of $/mps$.  Residuals to a classic, polynomial fit is shown in the top plot.  The center plot shows residuals to an interpolated wavelength solution.  The bottom plot shows residuals when a ThAr wavelength model generated by \name\ is interpolated onto LFC line positions.  The position of ThAr lines are over-plotted in black.}
\label{fig:resid2d}
\end{figure}

The residuals of the classic, polynomial wavelength solution is shown in the top plot of Figure \ref{fig:resid2d}.  There is a lot of vertical structure and some hints of a periodic, diagonal structure as well.  The residuals of the interpolation test for the same exposure is shown in the center plot of Figure \ref{fig:resid2d}.  There is no coherent structure here and smaller residuals.  This shows how the flexibility of an interpolated model can account for high-order detector defects, which emerged as structure in the residuals of the classic, smooth, polynomial-driven wavelength solution.  In the \expres\ data shown here, for example, the interpolated model accounted for \lz{detector effect STA will explain to us}.

The bottom plot of Figure \ref{fig:resid2d} shows the residuals when ThAr exposures, which have much fewer and less regular lines than LFC exposures, are run through \name\ and used to predict wavelengths for the completely independent LFC exposures taken during the same range of time.  Over-plotted in black are the positions of the ThAr lines.

@@

Though the interpolated wavelength solution returns lower, less-structured residuals than the polynomial wavelength solution when guided by LFC lines, we see in the bottom plot of Figure \ref{fig:resid2d} that an interpolated wavelength solution can return much worse results in the space between widely separated calibration lines.  Figure \ref{fig:waveResids} plots the residuals for a subset of one order for both a polynomial-based method and a PCHIP-based method guided by either ThAr lines or LFC lines.   The PCHIP model with ThAr lines (blue) curve returns huge residuals between two widely-spaced ThAr lines.  The classic, polynomial fit exhibits similar residuals in both amplitude and shape regardless of whether the set of ThAr lines or LFC lines are used.  An interpolated wavelength solution using LFC lines (orange curve) exhibits the lowest residuals.

The move to an interpolated wavelength solution is driven by the assumption that a high density of calibration lines allows for more freedom in the resultant wavelength solution.  This freedom allows the wavelength solution to more accurately ascribe wavelengths.  This flexibility, however, is no longer justified in the regime where there are large spaces between calibration lines, as is the case with a classic ThAr lamp.

\begin{figure}[t]
\centering
\includegraphics[width=0.5\textwidth]{Figures/lineDensity.png}
\caption{Comparison of polynomial-fit and interpolated wavelength solutions using either just ThAr lines or LFC lines.  For a subset of one order, the difference between the assigned and model wavelengths for each LFC line is plotted with respect to pixel.  The large residuals between two widely-spaced ThAr lines when implementing an interpolated wavelength solution with just ThAr lines (blue curve) extends out of frame.}
\label{fig:waveResids}
\end{figure}

\subsection{Impact on Radial Velocity Measurements}\label{sec:test-rv}
\lz{Awaiting latest extraction run.  Was supposed to take a weekend, but I'm pretty sure I heard God laugh when RP said that.}

\section{Choose Your Own Implementation} \label{sec:choices}
We have described and test only one implementation of \name.  Using PCA and an interpolated wavelength solution is a fairly trivial step towards a complete hierarchical and non-parametric wavelength model.  It is possible to upgrade both the denoising and wavelength solution steps to true models.

@@

The wavelength solution could also move past interpolation.  For example, a Gaussian process could be used that is constrained to ensure monotonicity.  Replacing each step with a model will allow for full, hierarchical Bayesian inference.  This means the uncertainty from wavelength calibration could be completely marginalized out.  Doing so, however, is only worth it if the wavelength calibration is a significant fraction of the error budget.

The implementation of \name\ presented here, using PCA for denoising and interpolating a wavelength solution, uses various global variables and methods that we believe are or are close to optimal for constructing a high-fidelity wavelength solution.  The following subsections will describe each choice and the associated decision-making process.

@@

We settled on a $K$ value of 6.  Figure \ref{fig:pcLfc} shows the first 9 principal components constructed using LFC lines.  There is clear structure in the first and second principal components.  \lz{The diagonal-banding structure seen in components 3, 5, and 6 we believe to be capturing detector fabrication defects. (Maybe, awaiting STA response)}  Components 3 and 4 show aberrant behavior on the edges of the two bluest echelle orders, where lower signal results in more variation in the line fits.  Later principal components become dominated by noise over coherent structure.

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{Figures/pcsLfc9.png}
%\includegraphics[width=\textwidth]{Figures/pcsLfc6.pdf}
\caption{The first nine principal components constructed using LFC lines.  Each line is plotted according to its echelle order and x-pixel and colored by the value of the principal component for that line.}
\label{fig:pcLfc}
\end{figure*}

In deciding a K value, we ran denoising tests, as described in Section \ref{sec:tests}, for K values spanning from 2 to 512.  One would expect the returned wavelengths to get increasingly more accurate with larger K until components that represent only noise are incorporated.  Residuals might then get worse before ultimately starting to get better again with large K, which marks te model over-fitting the data.  Though the returned RMS never gets worse, we find that the improvement plateaus between $K=6$ and $K=128$.  Comparisons of wavelengths returned by a $K=6$ model vs. a $K=32$ model show significant differences in less than 10 bluer lines, which are known to have greater error and variance in their measured line positions.  A visual inspection of the principal components, such as the one displayed in Figure \ref{fig:pcLfc}, additionally show most of the structure is being captured in the first six principal components with later components dominated by noise.

@@

The choice in interpolation method can help guide how many wavelength calibration images are truly needed.  It is unnecessary to take calibration images at times where the same information can be constructed by a well-chosen interpolation scheme.  For example, with the \expres\ data shown here, it is clear that nightly calibration images are needed, but for a linear trend, only two calibration images throughout the night are strictly required.

The amplitudes $a_{n,k}$ can also be interpolated with respect to any housekeeping data, not just time.  Best results will come from interpolating with respect to whichever data is most strongly correlated with the calibration state of the spectrograph.  For example, with \expres, which is passively temperature controlled, the returned amplitudes $a_{n,k}$ were extremely correlated with the enclosure temperature, suggesting it would also be possible to interpolate with respect to temperature.
